A Kochen-Specker system has at least 21 vertices\\ {\small (extended abstract)}

### 2014-05-06 main.tex
minor comments of anonymous reviewer #1

\subsection{Kochen-Specker graphs}
\label{sec:ksgraphs}
We follow \cite{aow11} and reduce the search for Kochen-Specker systems
to the search of a certain class of graphs.
First note that in a Kochen-Specker system we may replace a point with its
antipodal point.  They are both orthogonal to the same points and hence
the non-010-colorability is preserved.
Therefore, we may assume antipodal points are identified on the sphere.
That is: a Kochen-Specker system is a finite subset of the projective plane
that is not 010-colorable.
\begin{dfn}
Given a finite subset~$S$ of the projective plane
(or equivalently, of the northern hemisphere without equator).
Define its \keyword{orthogonality graph}~$G(S)$ as follows.
The vertices are the points of~$S$.
Two vertices are joined by an edge, if their corresponding points
are orthogonal.
\end{dfn}
\begin{dfn}
A graph~$G$ is called~\keyword{embeddable},
if it occurs as a subgraph of an orthogonality graph;
that is: if there is a finite subset~$S$ of the projective plane,
such that~$G \leq G(S)$.
\end{dfn}
% TODO add remark on subgraph definition?
\begin{dfn}
A graph is~\keyword{010-colorable}
if there is a~$\{0,1\}$-coloring, such that
\begin{enumerate}
\item
for each triangle there is exactly one vertex that is colored~$1$ and
\item
adjacent vertices are not both colored~$1$.
\end{enumerate}
\end{dfn}
\begin{dfn}
A \keyword{Kochen-Specker graph}
is a embeddable graph that is not 010-colorable.
\end{dfn}
It is an easy, but important, consequence of the definitions that:
\begin{fact}
    A finite subset~$S$ of the projective plane
    is a Kochen-Specker system,
    if and only if its orthogonality graph~$G(S)$
    is Kochen-Specker.
\end{fact}
To prove there is no Kochen-Specker system on~$17$ points,
it would be sufficient to enumerate all graphs on~$17$ vertices
and check these are not 010-colorable or not embeddable.
However, this is infeasible as there are
already~${\sim}10^{26}$ non-isomorphic
graphs on~$17$ points.
% TODO OEIS A000088
Luckily, we can restrict ourselves to certain classes of graphs.
\begin{prop}[\cite{aow11}]
    An embeddable graph does not contain a square.
\end{prop}
\begin{proof}
    Given two non antipodal points~$a\neq b$.
    Consider the points orthogonal to~$a$.
    This is a great circle.
    The points orthognal to~$b$ is a different great circle.
    They intersect in precisely two antipodal points.
    Hence, if~$c$ and~$d$ are both orthogonal to~$a$ and~$b$,
    then~$c$ and~$d$ are equivalent.
    Therefore, an embeddable graph cannot contain a square.
\end{proof}
The squarefreeness is a considerable restriction.  There are
only~${\sim}10^{10}$ non-isomorphic squarefree graphs on~$17$ vertices.
We can restrict ourselves to connected graphs.
% TODO OEIS A006786
\begin{prop}[\cite{aow11}]\label{prop:ks-conn}
    A minimal Kochen-Specker graph is connected.
\end{prop}
\begin{proof}
    Suppose~$G$ is a non-connected Kochen-Specker graph.
    Then one of its components is not 010-colorable.
    As a subgraph of an embeddable graph, is embeddable,
    this component is embeddable as well.
    Hence it is a smaller connected Kochen-Specker graph.
\end{proof}
The gain, however, is small.
There are only~${\sim}10^9$ non-isomorphic squarefree graphs on~$17$
vertices that are not connected.
In our computations, checking for connectedness
required more time than would be gained by reducing the number of graphs.

@@

We continue with a strengthening of Proposition~\ref{prop:ks-conn}.
\begin{prop}
A minimal Kochen-Specker graph is biconnected,
that is: removing any single edge leaves the graph connected.
\end{prop}
\begin{proof}
Suppose we have a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.

@@

Rotate the vectors of the embedding
along~$b$ such that~$b_2$ becomes orthogonal to~$a$.
It could that there are elements~$a'$ in~$A$
and~$b'$ in~$B$ that are rotated onto the same point.
Then consider rotations of all elements of~$B$ along~$a$.
There are finitely many angles for which two elements will overlap.
Hence, there is a rotation of all elements of~$B$ along~$a$
such that~$A$ and~$B$ do not overlap.
Note that all existing orthogonality relations are preserved.

@@

In this way, there were still a few (010-colorable) graphs of which
we could not decide embeddability.
By hand, we determined these graphs to be embeddable.
We addapted the algorithm, as to guess
for some assignments the position of one of the vectors.
If the corresponding
sentence turns out false, we know nothing.  However,
if the sentence is true, we know the graph is embeddable.

>>>>

\subsection{Kochen-Specker graphs}
\label{sec:ksgraphs}
We follow \cite{aow11} and reduce the search for Kochen-Specker systems
to the search of a certain class of graphs.
First note that in a Kochen-Specker system we may replace a point with its
antipodal point.  They are both orthogonal to the same points and hence
the non-010-colorability is preserved.
Therefore, we may assume antipodal points are identified on the sphere.
That is: a Kochen-Specker system is a finite subset of the projective plane
that is not 010-colorable.
\begin{dfn}
Given a finite subset~$S$ of the projective plane
(or equivalently, of the northern hemisphere without equator).
Define its \keyword{orthogonality graph}~$G(S)$ as follows.
The vertices are the points of~$S$.
Two vertices are joined by an edge, if their corresponding points
are orthogonal.
\end{dfn}
\begin{dfn}
A graph~$G$ is called~\keyword{embeddable},
if it occurs as a subgraph of an orthogonality graph;
that is: if there is a finite subset~$S$ of the projective plane,
such that~$G \leq G(S)$.
\end{dfn}
% TODO add remark on subgraph definition?
\begin{dfn}
A graph is~\keyword{010-colorable}
if there is a~$\{0,1\}$-coloring, such that
\begin{enumerate}
\item
for each triangle there is exactly one vertex that is colored~$1$ and
\item
adjacent vertices are not both colored~$1$.
\end{enumerate}
\end{dfn}
\begin{dfn}
A \keyword{Kochen-Specker graph}
is a embeddable graph that is not 010-colorable.
\end{dfn}
It is an easy, but important, consequence of the definitions that:
\begin{fact}
    A finite subset~$S$ of the projective plane
    is a Kochen-Specker system,
    if and only if its orthogonality graph~$G(S)$
    is Kochen-Specker.
\end{fact}
To prove there is no Kochen-Specker system on~$17$ points,
it would be sufficient to enumerate all graphs on~$17$ vertices
and check these are not 010-colorable or not embeddable.
However, this is infeasible as there are
already~${\sim}10^{26}$ non-isomorphic
graphs on~$17$ points.
% TODO OEIS A000088
Luckily, we can restrict ourselves to certain classes of graphs.
\begin{prop}[\cite{aow11}]
    An embeddable graph is squarefree.
    That is: it does not contain the square as a subgraph.
    \footnote{Some authors call a graph squarefree, if it does not
        contain the square as induced subgraph.
        For them the full graph on four nodes is squarefree.
        We follow~Weisstein\cite{sf-weisstein} and~Sloane\cite{sf-sloane} and
        call a graph squarefree if it does not
        contain the square as subgraph.
        For us the full graph on four nodes is not squarefree.}
\end{prop}
\begin{proof}
    Given two non antipodal points~$a\neq b$.
    Consider the points orthogonal to~$a$.
    This is a great circle.
    The points orthogonal to~$b$ is a different great circle.
    They intersect in precisely two antipodal points.
    Hence, if~$c$ and~$d$ are both orthogonal to~$a$ and~$b$,
    then~$c$ and~$d$ are equivalent.
    Therefore, an embeddable graph cannot contain a square.
\end{proof}
The squarefreeness is a considerable restriction.  There are
only~${\sim}10^{10}$ non-isomorphic squarefree graphs on~$17$ vertices.
We can restrict ourselves to connected graphs.
% TODO OEIS A006786
\begin{prop}[\cite{aow11}]\label{prop:ks-conn}
    A minimal Kochen-Specker graph is connected.
\end{prop}
\begin{proof}
    Suppose~$G$ is a non-connected Kochen-Specker graph.
    Then one of its components is not 010-colorable.
    As a subgraph of an embeddable graph, is embeddable,
    this component is embeddable as well.
    Hence it is a smaller connected Kochen-Specker graph.
\end{proof}
The gain, however, is small.
There are only~${\sim}10^9$ non-isomorphic squarefree graphs on~$17$
vertices that are not connected.
In our computations, checking for connectedness
required more time than would be gained by reducing the number of graphs.

@@

We continue with a strengthening of Proposition~\ref{prop:ks-conn}.
\begin{prop}
A minimal Kochen-Specker graph is biconnected,
that is: removing any single edge leaves the graph connected.
\end{prop}
\begin{proof}
Suppose we have a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.  Thus~$B(G)$ is a tree.

@@

Rotate the vectors of the embedding
along~$b$ such that~$b_2$ becomes orthogonal to~$a$.
It could be that there are elements~$a'$ in~$A$
and~$b'$ in~$B$ that are rotated onto the same point.
Then consider rotations of all elements of~$B$ along~$a$.
There are finitely many angles for which two elements will overlap.
Hence, there is a rotation of all elements of~$B$ along~$a$
such that~$A$ and~$B$ do not overlap.
Note that all existing orthogonality relations are preserved.

@@

In this way, there were still a few (010-colorable) graphs of which
we could not decide embeddability.
By hand, we determined these graphs to be embeddable.
We adapted the algorithm, as to guess
for some assignments the position of one of the vectors.
If the corresponding
sentence turns out false, we know nothing.  However,
if the sentence is true, we know the graph is embeddable.
